Shrinking Characteristics of Precision Matrix Estimators: An Illustration via Regression

### 2019-04-21 Manuscript.tex
Deploy MGallow/oral_manuscript to github.com/MGallow/oral_manuscript.git:gh-pages

\usepackage{booktabs}
\usepackage{amsthm}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother
\usepackage[]{natbib}
\bibliographystyle{apalike}

\title{Oral Manuscript}
\author{Matt Galloway}
\date{2018-06-07}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\theoremstyle{definition}
\newtheorem{example}{Example}[chapter]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[chapter]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{prerequisites}{%
\chapter{Prerequisites}\label{prerequisites}}

\hypertarget{intro}{%
\chapter{Introduction}\label{intro}}

This is the introduction for Matt's oral manuscript. (UPDATED 8)

\hypertarget{tutorial}{%
\chapter{Tutorial}\label{tutorial}}

>>>>

\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{fancyhdr} % Required for custom headers
\usepackage{lastpage} % Required to determine the last page for the footer
\usepackage{extramarks} % Required for headers and footers
\usepackage{graphicx}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother
\pagestyle{fancy}
% \pagestyle{empty}
\usepackage[left]{lineno} % line number package
\linenumbers % line number command
\usepackage[]{natbib}
\bibliographystyle{apalike}

\title{Shrinking Characteristics of Precision Matrix Estimators: An Illustration via Regression}
\author{Matt Galloway}
\date{2019-04-21}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\newtheorem{conjecture}{Conjecture}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\theoremstyle{definition}
\newtheorem{example}{Example}[chapter]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[chapter]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
\begin{document}
\maketitle

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{intro}{%
\chapter{Introduction}\label{intro}}

Inferential statistics consists of two primary components: estimation and prediction. Estimation refers to the process by which we infer values, properties and behavior about individual parameters in our model. Prediction, on the other hand, refers to the process by which we draw inference about an event that has not-yet occurred. The latter component is abundant in industry where the demand to predict stock prices, movies individuals prefer, and even to predict new friend connections in a network has sparked huge investments in research and given rise to research groups whose primary goal is pushing the bounds and developing new predictive models.

However, though most attention is directed towards prediction, these two components are not distinct. In fact, the success and reliability of many predictive models is contingent upon good, efficient parameter estimation. Many of these models require the estimation of a precision matrix - the inverse of the covariance matrix (frequently denoted as \(\Omega\)) - that establishes the interaction and covariance between random variables. For this reason, the last decade has seen an ever-expanding community devoted to precision matrix estimation.

Among this community of researchers is Professor Adam Rothman and Aaron Molstad, Ph.D.~whose research will be a focal point of this manuscript. The two have published work on indirect multivariate linear regression (\citet{molstad2016indirect}) and classification with matrix-valued predictors (\citet{molstad2018penalized}) but the focus of this manuscript is their 2017 paper titled \emph{Shrinking Characteristics of Precision Matrix Estimators} (\citet{molstad2017shrinking}). In it, they outline a framework to estimate \emph{characteristics} of precision matrices - a concept that exploits the fact that in many predictive models estimation of the precision matrix is only necessary through its product with another feature. They write in their manuscript that ``to fit many predictive models, only a characteristic of the population precision matrix needs to be estimated\ldots{} In binary linear discriminant analysis, the population precision matrix is needed for prediction only through the product of the precision matrix and the difference between the two conditional distribution mean vectors.'' The purpose of the research detailed here began with the desire to expand on this concept and to explore avenues that were mentioned but were not further investigated.

One of the research directions mentioned in the original paper was the application of their framework to regression. Utilizing the fact that the population regression coefficient matrix \(\beta \equiv \Omega_{x}\Sigma_{xy}\) (where \(\Sigma_{xy}\) is the cross-covariance matrix between the predictors, \(X\), and the responses, \(Y\), and \(\Omega_{x}\) is the precision matrix for \(X\)), their framework allows for the simultaneous estimation of \(\beta\) and \(\Omega_{x}\) with an embedded assumption potentially useful for superior prediction performance. In close communication and collaboration with Professor Rothman, we wanted to explore this research direction further. However, in order to build upon their work and contribute new material, there were a number of concepts that needed to be learned along the way and this document will follow that journey.

We will begin chapter two with a brief introduction to precision matrix estimation and the gaussian log-likelihood function. This section will mention popular estimation methods and algorithms but most discussion will be directed towards the ADMM algorithm. Discussion of the ADMM algorithm will be useful as we begin detailing the shrinking characteristics of precision matrix estimators framework (which may be referred to as SCPME), the so-called augmented ADMM algorithm, and later the framework's application to regression. Lastly, the document will end with two brief tutorials for the R packages \texttt{ADMMsigma} and \texttt{SCPME}. These packages were developed by myself to aid in simulation experiments and make it easier to branch into related research directions. Both packages have since been published on CRAN.

\hypertarget{notation-and-definitions}{%
\subsection{Notation and Definitions}\label{notation-and-definitions}}

For strictly positive integers \(n\) and \(p\), we will denote \(\mathbb{R}^{n \times p}\) as the class of real matrices with dimenson \(n \times p\). The class of real, symmetric matrices with dimension \(p \times p\) will be denoted as \(\mathbb{S}^{p}\) and \(\mathbb{S}^{p}_{+}\) if we further require the object to be positive definite. The sample size and dimension of the predictor vector in a given data set will most often be denoted as \(n\) and \(p\), respectively. If the dimension of the response vector exceeds one, we will denote it as \(r\).

Most matrices will take the form of either \(\Sigma\), the population covariance matrix, or \(\Omega\), the population precision matrix. Note that the precision matrix is simply the inverse of the covariance matrix (\(\Omega \equiv \Sigma^{-1}\)) and a subscript may be added to each if more than two random variables are considered in a problem (\(\Omega_{x}\)). A subscript star may also be added if the object is oracle - or known - a priori (\(\Omega_{*}\)). The oracle's estimator that optimizes a pre-specified objective function will be denoted with a hat (\(\hat{\Omega}\)).

There will be significant matrix algebra notation throughout the manuscript. The trace operator sums the diagonal elements of a matrix and will take the form \(tr\left(\cdot\right)\) and the \emph{exponential} trace operator will be denoted similarly as \(etr\left(\cdot\right)\). The vector operator, \(vec\left(\cdot\right)\), stacks the columns of a matrix into a column vector. The determinant of a matrix \(\mathbf{A}\) will be denoted as \(\left|\mathbf{A}\right|\) but may also take the form \(det\left(\mathbf{A}\right)\). The kronecker product of two matrices \(\mathbf{A}\) and \(\mathbf{B}\) will be denoted as \(\mathbf{A} \otimes \mathbf{B}\) and the element-wise product will be denoted as \(\mathbf{A} \circ \mathbf{B}\). Lastly, the Frobenius norm which sums the square of all entries in a matrix will be denoted as \(\left\|\mathbf{A}\right\|_{F}\) and we will define \(\left\|\mathbf{A}\right\|_{1} := \sum_{i, j}\left|\mathbf{A}_{ij}\right|\) where the \(i\)-\(j\)th element in matrix \(\mathbf{A}\) is typically denoted as \(\left(\mathbf{A}\right)_{ij}\) or simply \(\mathbf{A}_{ij}\).

\hypertarget{precision-matrix-estimation}{%
\chapter{Precision Matrix Estimation}\label{precision-matrix-estimation}}

\hypertarget{background}{%
\section{Background}\label{background}}

The foundation of much of the precision matrix estimation literature is the gaussian negative log-likelihood. Consider the case where we observe \(n\) independent, identically distributed (iid) copies of the random variable \(X\), where the \(i\)th observation \(X_{i} \in \mathbb{R}^{p}\) is normally distributed with mean, \(\mu\), and variance, \(\Omega^{-1}\). That is, \(X_{i}\) follows a \(p\)-dimensional normal distribution which is typically denoted as \(X_{i} \sim N_{p}\left( \mu, \Omega^{-1} \right)\). By definition, this multivariate formulation implies the probability distribution function, \(f\), is of the form

\begin{equation} 
  f\left(X_{i}; \mu, \Omega\right) = (2\pi)^{-p/2}\left| \Omega \right|^{1/2}\exp\left[ -\frac{1}{2}\left( X_{i} - \mu \right)'\Omega\left( X_{i} - \mu \right) \right] 
\label{eq:normallikone}\notag
\end{equation}

Furthermore, because we assume that each observation is independent, the probability distribution function for all \(n\) observations \(X_{1}, ..., X_{n}\) is equal to

\begin{equation} 
\begin{split}
  f\left(X_{1}, ..., X_{n}; \mu, \Omega\right) &= \prod_{i = 1}^{n}(2\pi)^{-p/2}\left| \Omega \right|^{1/2}\exp\left[ -\frac{1}{2}\left( X_{i} - \mu \right)'\Omega\left( X_{i} - \mu \right) \right] \\
  &= (2\pi)^{-np/2}\left| \Omega \right|^{n/2}\mbox{etr}\left[ -\frac{1}{2}\sum_{i = 1}^{n}\left( X_{i} - \mu \right)\left( X_{i} - \mu \right)'\Omega \right]
\end{split}
\label{eq:normallik}\notag
\end{equation}

Therefore, the gaussian log-likelihood, \(l\), for \(\mu\) and \(\Omega\) given \(X = (X_{1}, .., X_{n})\) can be written as

\begin{equation}
  l(\mu, \Omega | X) = constant + \frac{n}{2}\log\left| \Omega \right| - tr\left[ \frac{1}{2}\sum_{i = 1}^{n}\left(X_{i} - \mu \right)\left(X_{i} - \mu \right)'\Omega \right]
\label{eq:normalloglik}\notag
\end{equation}

The estimator for \(\mu\) that maximizes the log-likelihood is \(\hat{\mu}^{mle} = \bar{X} \equiv \sum_{i = 1}^{n}X_{i}/n\), so that the partially maximized gaussian log-likelihood function for \(\Omega\) is

\begin{equation}
\begin{split}
  l(\Omega | X) &= \frac{n}{2}\log\left| \Omega \right| - tr\left[ \frac{1}{2}\sum_{i = 1}^{n}\left(X_{i} - \bar{X} \right)\left(X_{i} - \bar{X} \right)'\Omega \right] \\
  &= \frac{n}{2}\log\left| \Omega \right| - \frac{n}{2}tr\left( S\Omega \right)
\label{eq:partialmaxloglik}\notag
\end{split}
\end{equation}

where \(S = \sum_{i = 1}^{n}\left(X_{i} - \bar{X}\right)\left(X_{i} - \bar{X}\right)'/n\) is the usual sample estimator for the population covariance matrix, \(\Sigma\). In addition to \(\mu\), one could also derive the maximum likelihood estimator for \(\Omega\). By setting the gradient of the partially maximized log-likelihood equal to zero, one could show that

\begin{equation}
\begin{split}
\hat{\Omega}^{mle} &= \arg\max_{\Omega \in S_{+}^{p}}\left\{ \frac{n}{2}\log\left| \Omega \right| - \frac{n}{2}tr\left(S\Omega \right) \right\} \\
  &= \arg\min_{\Omega \in S_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega\right| \right\} \\
  &= S^{-1}
\end{split}
\label{eq:omegamle}\notag
\end{equation}

so that the MLE for \(\Omega\), when it exists, is \(\hat{\Omega}^{mle} = S^{-1} = \left[\sum_{i = 1}^{n}\left(X_{i} - \bar{X}\right)\left(X_{i} - \bar{X}\right)'/n \right]^{-1}\). The reality, however, is that this object does \emph{not} always exist. In settings where the number of observations is exceeded by the number of features in a sample, the sample covariance matrix is rank deficient and no longer invertible. For this reason, many papers in the last decade have proposed \emph{shrinkage estimators} of the population precision matrix similar to the shrinkage estimators in regression settings. That is, instead of minimizing solely the negative log-likelihood function, researchers have proposed minimizing the gaussian log-likelihood \emph{plus} a penalty term, \(P\), where \(P\) is often a function of the precision matrix.

\begin{equation}
\hat{\Omega} = \arg\min_{\Omega \in S_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega \right| + P\left( \Omega \right) \right\}
\label{eq:omegapen}
\end{equation}

The penalties that have been proposed for precision matrix estimation are typically a variation of the ridge penalty \(P\left(\Omega \right) = \lambda\|\Omega \|_{F}^{2}/2\) or the lasso penalty \(P\left(\Omega \right) = \lambda\left\|\Omega\right\|_{1}\) (here \(\lambda\) is a tuning parameter). The authors, \citet{yuan2007model}, initially proposed the lasso-penalized gaussian log-likelihood defined as

\begin{equation}
\hat{\Omega} = \arg\min_{\Omega \in S_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega \right| + \lambda\sum_{i \neq j}\left|\Omega_{ij}\right| \right\}
\label{eq:omegapenlasso}
\end{equation}

so as not to penalize the diagonal elements of the precision matrix estimate. Other papers published on the lasso-penalized gaussian likelihood precision matrix estimator include \citet{rothman2008sparse} and \citet{friedman2008sparse}. In addition, many efficient algorithms have been proposed to solve for \(\hat{\Omega}\), however, the most popular method is the graphical lasso algorithm (glasso) introduced by \citet{friedman2008sparse}. Their method utilizes an iterative block-wise coordinate descent algorithm that builds upon the coordinate descent algorithm used in lasso-penalized regression.

Non-lasso, non-convex penalties were considered in \citet{lam2009sparsistency} and \citet{fan2009network} and other papers considered penalizations like the Frobenius norm (\citet{rothman2014existence}; \citet{witten2009covariance}; \citet{price2015ridge}). In fact, the latter two papers show that the resulting minimizer can be solved in closed-form - which will be discussed later in this manuscript. However, the penalty explored through the remainder of this chapter is not a lasso penality nor a ridge penalty but, in fact, a convex combination of the two known as the \emph{elastic-net} penalty:

\begin{equation}
P\left( \Omega \right) = \lambda\left[\frac{1 - \alpha}{2}\left\| \Omega \right\|_{F}^{2} + \alpha\left\| \Omega \right\|_{1} \right]
\label{eq:penelastic}\notag
\end{equation}

with additional tuning parameter \(0 \leq \alpha \leq 1\). Clearly, when \(\alpha = 0\) this penalty reduces to a ridge penalty and when \(\alpha = 1\) it reduces to a lasso penalty. Originally proposed in \citet{zou2005regularization} in the context of regression, this penalty has since been popularized and is used in the penalized regression R package \texttt{glmnet}. This penalty allows for additional flexibility but, despite this, no published work to our knowledge has explored the elastic-net penalty in the context of precision matrix estimation. We will show how to solve the following optimization problem in the next section using the ADMM algorithm.

\begin{equation}
\hat{\Omega} = \arg\min_{\Omega \in S_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega \right| + \lambda\left[\frac{1 - \alpha}{2}\left\| \Omega \right|_{F}^{2} + \alpha\left\| \Omega \right\|_{1} \right] \right\}
\label{eq:optimelastic}
\end{equation}

ADMM stands for alternating direction method of multipliers. The algorithm was largely popularized by Stephen Boyd and his fellow authors in the book \emph{Distributed Optimization and Statistical Learning via the Alternating Direction Method of Multipliers} \citep{boyd2011distributed}. As the authors state in the text, the ``ADMM is an algorithm that is intended to blend the decomposability of dual ascent with the superior convergence properties of the method of multipliers.'' By closely following Boyd's descriptions and guidance in the published text, we will show in this section that the ADMM algorithm is particularly well-suited to solve the penalized log-likelihood optimization problem we are interested in here.

In general, the ADMM algorithm supposes that we want to solve an optimization problem of the form

\begin{equation}
\begin{split}
  \mbox{minimize } f(x) + g(z) \\
  \mbox{subject to } Ax + Bz = c
\end{split}
\label{eq:boyd}\notag
\end{equation}

where we can assume here that \(x \in \mathbb{R}^{n}, z \in \mathbb{R}^{m}, A \in \mathbb{R}^{p \times n}, B \in \mathbb{R}^{p \times m}\), \(c \in \mathbb{R}^{p}\), and \(f\) and \(g\) are convex functions. In order to find the pair \((x^{*}, z^{*})\) that achieves the infimum, the ADMM algorithm uses an \emph{augmented lagrangian}, \(L\), which \citet{boyd2011distributed} define as

\begin{equation}
L_{\rho}(x, z, y) = f(x) + g(z) + y'(Ax + Bz - c) + \frac{\rho}{2}\left\| Ax + Bz - c \right\|_{2}^{2}
\label{eq:auglagrange}\notag
\end{equation}

In this formulation, \(y \in \mathbb{R}^{p}\) is called the lagrange multiplier and \(\rho > 0\) is some scalar that acts as the step size for the algorithm. Note that any infimum under the augmented lagrangian is equivalent to the infimum of the traditional lagrangian since any feasible point \((x, z)\) must satisfy the constraint \(\rho\left\| Ax + Bz - c \right\|_{2}^{2}/2 = 0\). \citet{boyd2011distributed} show that using the ADMM algorithm the infimum will be approached under the following repeated iterations:

\begin{equation}
\begin{split}
  x^{k + 1} &= \arg\min_{x \in \mathbb{R}^{n}}L_{\rho}(x, z^{k}, y^{k}) \\
  z^{k + 1} &= \arg\min_{z \in \mathbb{R}^{m}}L_{\rho}(x^{k + 1}, z, y^{k}) \\
  y^{k + 1} &= y^{k} + \rho(Ax^{k + 1} + Bz^{k + 1} - c)
\end{split}  
\label{eq:admmalgo}\notag
\end{equation}

where the superscript, \(k\), denotes the number of iterations. Conveniently, this general algorithm can be coerced into a format useful in precision matrix estimation. Suppose we let \(f\) be equal to the non-penalized gaussian log-likelihood, \(g\) equal to the elastic-net penalty, \(P\left( \Omega \right)\), and we use the constraint that \(\Omega \in \mathbb{S}_{+}^{p}\) must be equal to some matrix \(Z \in \mathbb{R}^{p \times p}\), then the augmented lagrangian in the context of precision matrix estimation is of the form

\begin{equation}
L_{\rho}(\Omega, Z, \Lambda) = f\left(\Omega\right) + g\left(Z\right) + tr\left[\Lambda\left(\Omega - Z\right)\right] + \frac{\rho}{2}\left\|\Omega - Z\right\|_{F}^{2}
\label{eq:auglagrange2}\notag
\end{equation}

where \(\Lambda\) takes the role of \(y\) as the lagrange multiplier. The ADMM algorithm now consists of the following repeated iterations:

\begin{equation}
\begin{split}
  \Omega^{k + 1} &= \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega\right| + tr\left[\Lambda^{k}\left(\Omega - Z^{k}\right)\right] + \frac{\rho}{2}\left\| \Omega - Z^{k} \right\|_{F}^{2} \right\} \\
  Z^{k + 1} &= \arg\min_{Z \in \mathbb{S}^{p}}\left\{ \lambda\left[ \frac{1 - \alpha}{2}\left\| Z \right\|_{F}^{2} + \alpha\left\| Z \right\|_{1} \right] + tr\left[\Lambda^{k}\left(\Omega^{k + 1} - Z\right)\right] + \frac{\rho}{2}\left\| \Omega^{k + 1} - Z \right\|_{F}^{2} \right\} \\
  \Lambda^{k + 1} &= \Lambda^{k} + \rho\left( \Omega^{k + 1} - Z^{k + 1} \right)
\label{eq:ADMMorig}
\end{split}
\end{equation}

Furthermore, it turns out that each step in this algorithm can be solved efficiently in closed-form. The full details of each can be found in the appendix \ref{proofomegaalgo} but the following theorem provides the simplified steps in the algorithm.

\BeginKnitrBlock{theorem}[ADMM Algorithm for Elastic-Net Penalized Precision Matrix Estimation]
\protect\hypertarget{thm:unnamed-chunk-2}{}{\label{thm:unnamed-chunk-2} \iffalse (ADMM Algorithm for Elastic-Net Penalized Precision Matrix Estimation) \fi{} }
Define the soft-thresholding function as \(\mbox{soft}(a, b) = \mbox{sign}(a)(\left| a \right| - b)_{+}\) and \(S\) as the sample covariance matrix. Set \(k = 0\) and initialize \(Z^{0}, \Lambda^{0}\), and \(\rho\). Repeat steps 1-3 until convergence:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Decompose \(S + \Lambda^{k} - \rho Z^{k} = VQV'\) via spectral decomposition\footnote{Proof of \eqref{eq:omegaalgo} in section \ref{proofomegaalgo}.}. Then
\end{enumerate}

\begin{equation}
\Omega^{k + 1} = \frac{1}{2\rho}V\left[ -Q + \left( Q^{2} + 4\rho I_{p} \right)^{1/2} \right]V'
\label{eq:omegaalgo}
\end{equation}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Elementwise soft-thresholding for all \(i = 1,..., p\) and \(j = 1,..., p\)\footnote{Proof of \eqref{eq:ZZalgo} in section \ref{proofZZalgo}.}.
\end{enumerate}

\begin{equation}
\begin{split}
Z_{ij}^{k + 1} &= \frac{1}{\lambda(1 - \alpha) + \rho}\mbox{sign}\left(\rho\Omega_{ij}^{k + 1} + \Lambda_{ij}^{k}\right)\left( \left| \rho\Omega_{ij}^{k + 1} + \Lambda_{ij}^{k} \right| - \lambda\alpha \right)_{+} \\
&= \frac{1}{\lambda(1 - \alpha) + \rho}\mbox{soft}\left(\rho\Omega_{ij}^{k + 1} + \Lambda_{ij}^{k}, \lambda\alpha\right)
\label{eq:ZZalgo}
\end{split}
\end{equation}

\begin{equation}
\Lambda^{k + 1} = \Lambda^{k} + \rho\left( \Omega^{k + 1} - Z^{k + 1} \right)
\label{eq:lamalgo}\notag
\end{equation}
\EndKnitrBlock{theorem}

\hypertarget{scaled-form-admm}{%
\subsection{Scaled-Form ADMM}\label{scaled-form-admm}}

Another popular, alternative form of the ADMM algorithm can be used when scaling the dual variable (\(\Lambda^{k}\)) which we will briefly mention in the context of precision matrix estimation here.

Define \(R^{k} = \Omega - Z^{k}\) and \(U^{k} = \Lambda^{k}/\rho\) then

\begin{equation}
\begin{split}
  tr\left[ \Lambda^{k}\left( \Omega - Z^{k} \right) \right] + \frac{\rho}{2}\left\| \Omega - Z^{k} \right\|_{F}^{2} &= tr\left[ \Lambda^{k}R^{k} \right] + \frac{\rho}{2}\left\| R^{k} \right\|_{F}^{2} \\
  &= \frac{\rho}{2}\left\| R^{k} + \Lambda^{k}/\rho \right\|_{F}^{2} - \frac{\rho}{2}\left\| \Lambda^{k}/\rho \right\|_{F}^{2} \\
  &= \frac{\rho}{2}\left\| R^{k} + U^{k} \right\|_{F}^{2} - \frac{\rho}{2}\left\| U^{k} \right\|_{F}^{2}
\label{eq:scaled}\notag
\end{split}
\end{equation}

Therefore, a scaled-form ADMM algorithm can now be written as

\begin{equation}
\begin{split}
  \Omega^{k + 1} &= \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega\right| + \frac{\rho}{2}\left\| \Omega - Z^{k} + U^{k} \right\|_{F}^{2} \right\} \\
  Z^{k + 1} &= \arg\min_{Z \in \mathbb{S}^{p}}\left\{ \lambda\left[ \frac{1 - \alpha}{2}\left\| Z \right\|_{F}^{2} + \alpha\left\| Z \right\|_{1} \right] + \frac{\rho}{2}\left\| \Omega^{k + 1} - Z + U^{k} \right\|_{F}^{2} \right\} \\
  U^{k + 1} &= U^{k} + \Omega^{k + 1} - Z^{k + 1}
\label{eq:ADMMscaled}
\end{split}
\end{equation}

Note that there are limitations to using this method. Because the dual variable is scaled by \(\rho\) (the step size), this form limits one to using a constant step size for all \(k\) steps if no further adjustments are made to \(U^{k}\).

\hypertarget{ADMMstop}{%
\subsection{Stopping Criterion}\label{ADMMstop}}

There are three optimality conditions for the ADMM algorithm that we can use to inform the convergence and stopping criterion. These general conditions were outlined in \citet{boyd2011distributed} but here we cater them to precision matrix estimation. The first condition is the primal optimality condition \(\Omega^{k + 1} - Z^{k + 1} = 0\) and the other two conditions are the dual conditions \(0 \in \partial f\left(\Omega^{k + 1}\right) + \Lambda^{k + 1}\) and \(0 \in \partial g\left(Z^{k + 1}\right) - \Lambda^{k + 1}\). The first dual optimality condition is a result of taking the sub-differential of the lagrangian (non-augmented) with respect to \(\Omega^{k + 1}\) and the second is a result of taking the sub-differential of the lagrangian with respect to \(Z^{k + 1}\). Note that in the first condition we must honor the symmetric constraint in \(\Omega\) but the second condition does not require it.

If we define the left-hand side of the primal optimality condition as the \emph{primal residual} \(r^{k + 1} = \Omega^{k + 1} - Z^{k + 1}\), then at convergence we must require that \(r^{k + 1} \approx 0\). Likewise, if we define the \emph{dual residual} \(s^{k + 1} = \rho\left( Z^{k + 1} - Z^{k} \right)\), we must also require that \(s^{k + 1} \approx 0\) for proper convergence. This dual residual is the direct result of the fact that \(\Omega^{k + 1}\) is the minimizer of the augmented lagragian\footnote{Proof in section \ref{proofdualresidual}.} so that \(0 \in \partial L_{p}\left( \Omega, Z^{k}, \Lambda^{k} \right)\) and consequently \(0 \in \rho\left( Z^{k + 1} - Z^{k} \right)\)\footnote{Note that the second dual optimality condition \(0 \in \partial g\left(Z^{k + 1}\right) - \Lambda^{k + 1}\) is always satisfied. More details can be found in section \ref{proofdualopt}.}.

Combining these three optimality conditions, Boyd suggests a stopping criterion similar to \(\epsilon^{pri} \leq \left\| r^{k + 1} \right\|_{F}\) and \(\epsilon^{dual} \leq \left\| s^{k + 1} \right\|_{F}\) where \(\epsilon^{rel} = \epsilon^{abs} = 10^{-3}\) and
\begin{equation}
\begin{split}
  \epsilon^{pri} &= p\epsilon^{abs} + \epsilon^{rel}\max\left\{ \left\| \Omega^{k + 1} \right\|_{F}, \left\| Z^{k + 1} \right\|_{F} \right\} \\
  \epsilon^{dual} &= p\epsilon^{abs} + \epsilon^{rel}\left\| \Lambda^{k + 1} \right\|_{F}
\label{eq:admmstopping}\notag
\end{split}
\end{equation}

\hypertarget{simulations}{%
\section{Simulations}\label{simulations}}

As a proof-of-concept that the elastic-net penalty in the context of precision matrix estimation can provide useful results and that the ADMM algorithm used in this process works, this section offers a short simulation. For the simulation, we generated data from multiple, unique precision matrices with various oracle structures. For each data-generating procedure, the algorithm was run with a 5-fold cross validation to tune parameters \(\lambda\) and \(\alpha\). After 20 replications, the cross validation errors were totalled and the optimal tuning parameters were selected (results are in the top half of the figures). These results were then compared with the Kullback Leibler (KL) losses between the estimated matrices and the oracle matrices (results are in the bottom half of the figures).

The first figure shows the results when the data was generated from a multivariate normal distribution with mean equal to zero and a tri-diagonal oracle precision matrix. This oracle matrix was first generated as \(\left(S_{ij}\right) = 0.7^{\left|i - j \right|}\) for \(i,j = 1,..., p\) and then inverted. The results show that because the oracle precision matrix is sparse, the algorithm correctly chooses a sparse solution with \(\alpha = 1\) - indicating a lasso penalty.

The second figure shows the results when the data was generated from a multivariate normal distribution with mean equal to zero and a \emph{dense} oracle precision matrix (non-sparse). Here, we randomly generated an orthogonal basis, set all eigen values equal to 1000, and then combined the matrices using QR decomposition. Interestingly, we find that the optimal \(\alpha\) in this case is 0.6 which closely matches the optimal result based on the KL loss. This shows that there are cases where an elastic-net penalty can provide useful results and that using only a lasso penalty may unnecessarily restrict our penalized estimation.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth,]{images/repsKLtridiag_N50_P100}

\caption{The oracle precision matrices were tri-diagonal with dimension p = 100 and the data was generated with a sample size of n = 50. The cross validation errors are in the top figure and the KL losses between the estimated matrices and the oracle matrices are shown in the bottom figure. The optimal tuning parameter pair for each heatmap was found to be log10(lam) = -0.9 and alpha = 1. Note that brighter areas signify smaller losses.}\label{fig:admmsim5}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth,]{images/repsKLdenseQR_N50_P100}

\caption{The oracle precision matrices were dense with dimension p = 100 and the data was generated with a sample size of n = 50. The cross validation errors are in the top figure and the KL losses between the estimated matrices and the oracle matrices are shown in the bottom figure. The optimal tuning parameter pair for the cross validation errors was found to be log10(lam) = -0.4 and alpha = 0.6 and log10(lam) = -0.5 and alpha = 0.7 for the KL losses. Note that brighter areas signify smaller losses.}\label{fig:admmsim3}
\end{figure}

\hypertarget{scpme}{%
\chapter{SCPME}\label{scpme}}

In their 2017 paper, titled \emph{Shrinking Characteristics of Precision Matrix Estimators}, Aaron Molstad, Ph.D.~and Professor Adam Rothman outline a framework to estimate \emph{characteristics} of precision matrices. This concept, inspired by others like \citet{cai2011direct}, \citet{fan2012road}, and \citet{mai2012direct}, exploits the fact that in many predictive models estimation of the precision matrix is only necessary through its product with another feature, such as a mean vector. The example they offer in \citet{molstad2017shrinking} is in the context of Fisher's linear discriminant analysis model. If a response vector \(Y\) is categorical such that \(Y\) can take values in \(\left\{1, ..., J\right\}\), then the linear discrimnant analysis model assumes that the design matrix \(X\) conditional on the response vector \(Y\) is normally distributed:

\begin{equation}
X | Y = j \sim N_{p}\left(\mu_{j}, \Omega^{-1}\right)
\label{eq:lda}
\end{equation}

for each \(j = 1, ..., J\). One can see from this formulation that clearly an estimation of both \(\mu\) and \(\Omega\) are required for this model. However, they note that if prediction is the primary concern, then for a given observation \(X_{i}\) only the \emph{characteristic} \(\Omega\left(\mu_{l} - \mu_{m}\right)\) is needed to discern between response categories \(l\) and \(m\). In other words, prediction only requires the characteristic \(\Omega\left(\mu_{l} - \mu_{m}\right)\) for each \(l, m \in \left\{1, ..., J\right\}\) and does \emph{not} require estimation of the full precision matrix \(\Omega\). \citet{cai2011direct} were among the first authors to propose estimating this characteristic directly but the interesting facet that distinguishes Molstad and Rothman's approach is that their framework simultaneously fits the model in \eqref{eq:lda} and performs variable selection.

The general framework has applications outside of linear discriminant analysis and we will be exploring a regression application in later sections, but first we will outline their approach. The penalty proposed by \citet{molstad2017shrinking} is of the form

\begin{equation}
P\left(\Omega\right) = \lambda\left\| A\Omega B - C \right\|_{1}
\label{eq:pen2}
\end{equation}

where \(A \in \mathbb{R}^{m \times p}, B \in \mathbb{R}^{p \times q}, \mbox{ and } C \in \mathbb{R}^{m \times q}\) are matrices that are assumed to be known and specified so that solving the full penalized gaussian negative log-likelihood for \(\Omega\) results in solving

\begin{equation}
\hat{\Omega} = \arg\min_{\Omega \in S_{+}^{p}}\left\{ tr\left(S\Omega\right) - \log\left|\Omega \right| + \lambda\left\| A\Omega B - C \right\|_{1} \right\}
\label{eq:omegaloglik2}
\end{equation}

A similar estimator was proposed by \citet{dalal2017sparse} when \(C = 0\) but here we do not require it. This form of penalty is particularly useful because it is extremely general. Note that by letting matrices \(A = I_{p}, B = I_{p}, \mbox{ and } C = 0\), this penalty reduces to a lasso penalty - but clearly this form allows for much more creative penalties and \(A, B, \mbox{ and } C\) can be constructed so that we penalize the sum, absolute value of \emph{many} characteristics of the precision matrix \(\Omega\). We will explore how to solve for \(\hat{\Omega}\) in \eqref{eq:omegaloglik2} in the next section.

\hypertarget{augmented-admm-algorithm}{%
\section{Augmented ADMM Algorithm}\label{augmented-admm-algorithm}}

Solving for \(\hat{\Omega}\) in \eqref{eq:omegaloglik2} uses what we are going to call an \emph{augmented ADMM algorithm}. Molstad and Rothman do not offer a name for this specific algorithm but it leverages the majorize-minimize principle in one of the steps in the algorithm - augmenting the original ADMM algorithm discussed in the previous chapter. Within the context of the proposed penalty, the original ADMM algorithm for precision matrix estimation would consist of iterating over the following three steps:

\begin{align}
  \Omega^{k + 1} &= \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}L_{\rho}(\Omega, Z^{k}, \Lambda^{k}) \\
  Z^{k + 1} &= \arg\min_{Z \in \mathbb{R}^{n \times r}}L_{\rho}(\Omega^{k + 1}, Z, \Lambda^{k}) \\
  \Lambda^{k + 1} &= \Lambda^{k} + \rho\left(A\Omega^{k + 1}B - Z^{k + 1} - C \right)
\end{align}
